---
title: "The Last Giants"
subtitle: "A data-driven safari into the savannah"
author: "Visual Virtuosos"
format: html
editor: visual
---

```{r}

if (!require("pacman")) 
  install.packages("pacman")

pacman::p_load(knitr,tidyverse,lubridate,here,dplyr,ggplot2,ggmap,maps,rnaturalearth,rnaturalearthdata,sf,plotly,caret,glmnet,steamgraph)
```

## Abstract

African elephants face severe threats from poaching and the loss of their natural habitats. Despite global attention towards their conservation, there is considerable uncertainty surrounding the population sizes and trends of elephants across much of Africa. Timely and accurate data on elephant populations is crucial for effective conservation efforts.

A continent-wide survey conducted by the `National Library of Medicine` revealed a stark decline in African Savannah Elephants. Africa may have held over 20 million elephants before European colonization and 1 million as recently as the 1970s. The survey estimated a population of 352,271 savannah elephants across study sites in 18 countries, representing approximately 93% of all savannah elephants in those areas. Alarmingly, regions with historical data indicated a significant decline, estimating a decrease of 144,000 elephants between 2007 and 2014. Moreover, current data suggests a continent-wide decline of 8% per year in elephant populations, primarily attributed to poaching. During the 1970s and 1980s, a surge in poaching significantly reduced elephant populations in numerous regions. Moreover, a resurgence of poaching incidents since approximately 2005 has resulted in an alarming annual death toll, estimated at around 30,000 elephants recently.

This project aims to examine how elephant populations across Africa changed between 1995 and 2015. It involves using graphs and maps to show how the number of elephants varied over these years. To understand what might happen in the future, the project will use different mathematical methods to guess how many elephants there might be from 2016 to 2023.

Moreover, this study will also look into whether there's a connection between how many elephants there are and the trade of ivory. By comparing data about how many elephants there are with information about ivory trade, the project will try to find out if ivory trade affects the number of elephants.

The main goals are to see how elephant numbers changed over time, predict how many might exist in the coming years, and understand if the ivory trade has any impact on elephant populations. This research aims to help conservationists better understand and protect these majestic animals.

## Introduction

The African elephant(Loxodonta africana) aka The Gentle Giants, a keystone species crucial to the continent's ecosystem, has faced significant population challenges over the decades due to various factors such as poaching, habitat loss, and human-wildlife conflicts. Understanding the historical trends in elephant populations is fundamental in formulating effective conservation strategies and policies. This project aims to comprehensively analyze the population dynamics of African elephants spanning from 1950 to 2015 by leveraging robust data visualization techniques and advanced statistical methods.

Using datasets sourced from reputable conservation organizations and wildlife databases, the project begins by visually mapping the fluctuations in elephant populations across different regions of Africa over the studied period. Subsequently, employing multiple regression methods, including but not limited to linear regression and time series analysis, the study intends to forecast the elephant population from 2016 to 2023 based on historical data trends.

Moreover, in an effort to comprehend the potential correlation between the illegal ivory trade and elephant populations, the project incorporates ivory trade data alongside elephant population data-set. Through correlation analysis and visual representations, the study aims to discern any discernible patterns or connections between these variables, shedding light on whether the ivory trade has had an impact on the fluctuating populations of African elephants.

By combining data visualization techniques, statistical analyses, and exploration of the complex interplay between elephant populations and the ivory trade, this project endeavors to contribute valuable insights into the dynamics of African elephant conservation and the factors influencing their populations. Ultimately, the findings may aid policymakers, conservationists, and wildlife experts in formulating more targeted and effective measures to safeguard these majestic creatures and their habitats.

## Question 1:- How has elephant population varied over the years across several African countries?

### Introduction

Elephants are the gentle giants and here we want to know about the trends in their population and to find out whether its declining or is at safe level, and if there is a decrease in their population then to find out the reason the cause. Understanding the population dynamics of African elephants is pivotal for conservation efforts and ecological balance. This will help us in safeguarding their future and to show to next many generations the gentle giants of this planet and how marvelous they are.

The dataset under examination spans from 1995 to 2014, presenting a valuable opportunity to predict and comprehend the trends in elephant populations for the subsequent years i.e. 2015 to 2023. This study aims to investigate the variations in elephants population across several African countries during the period, leveraging a multi-faceted analytical approach including data pre-processing, explanatory analysis, predictive modelling and country-level analysis

### Approach

###Here we are going to use the data-set about the elephants we have and would like to make some visualizations to represent and analyze the data in an appropriate manner. We are going to utilize these variables from the data `year`, `country` and `estimate` - We are going to make a line plot to find out the basic trend and continue to process to maybe plot the line plots for individual country. The data has 1995 to 2014 data, which we will utilize to find the trends and then use the present data to predict the data between 2015 to 2023 using multiple regression models.###

-   Data Pre-processing:

    -   

-   Data Exploration

    -   

-   Data Prediction

    -   

-   Model Evaluation

    -   

-   Country Level Analysis

    -   

-   Shiny App development

    -   

### Analysis

-   Data Pre-processing:

    -   

-   Data Exploration

    -   

-   Data Prediction

    -   

-   Model Evaluation

    -   

-   Country Level Analysis

    -   

-   Shiny App development

    -   

## Question 2:- What is the trend of ivory trade in Africa?

### Introduction

The illegal trade of ivory has remained a critical concern, posing significant threats to the African elephant population and biodiversity conservation. This analysis aims to study ivory trade in Africa using available historical data. The dataset covers past ivory trade information and seeks to predict trends from 2020 to 2023. The main goal is to reveal how ivory trade has changed over time by using careful data processing, exploring data, making predictions and creating clear visualizations.

### Approach

-   Data Pre-processing:

    -   

-   Data Exploration

    -   

-   Data Prediction

    -   

-   Model Evaluation

    -   

-   Country Level Analysis

    -   

-   Shiny App development

## Plots

```{r}

ivory_exports_data <- read_csv(here("data","cites_ivory_export.csv"))
```

```{r}

# Count the number of instances for each year
export_count_per_year <- ivory_exports_data |>
  group_by(Year) |>
  summarise(export_count = n())

# Assuming `count_per_year` is your dataset
export_count_per_year <- export_count_per_year |>
  filter(Year > 2000)

export_count_per_year

# Create a bar plot
ggplot(export_count_per_year, aes(x = as.factor(Year), y = export_count)) +
  geom_bar(stat = "identity", fill = "grey", color = "black") +
  labs(
    x = "Year",
    y = "Number of exports",
    title = "Ivory exports in Africa from 1990 to 2022",
    caption = "Source: CITES (Convention of International Trade 
    in Endangered Species of Wild Fauna and Flora)"
  ) +
  theme_minimal()
```
